Last glacial cycle glacier erosion potential in the Alps

### 2021-05-23 articles/erosion/response-referee1.md

> I think the study could also well contribute to the discussion on the
> elevation distribution of the cumulative glacial erosion over several cycles
> in the Alps (high elevations vs low elevation, e.g. Valla et al., 2011). A
> plot showing the integrated glacial erosion potential over (e.g. 100 m)
> elevation bins would easy to do and eventually an enlightening supplement
> summing up Fig. 5a.

TODO: try and compute cumulative erosion by elevation bins.

> I would also recommend to make some regional statements on the distribution
> of the glacial erosion potential (Fig. 2). It is quite obvious that some
> tectonic massifs can readily be discriminated, e.g. the Tauern Window, Oetzal
> Crystallin complex, the Aare Massif, Mt. Rosa - Gran Paradiso and Pelvoux
> Massif. Is it the steepness of these low erodible crystallin massifs making
> the erosional potential appearing strikingly high? I think this should/could
> be discussed..

More accurate modelling of small-scale glaciers would be needed to fully
address this point. Steep topography though is likely to be part of the
explanation, and a more substantial discussion of the high erosion potential
values in these areas and the model limitations that hinder interpretation was
added in the "age of the glacial landscape" subsection.

### 2021-05-23 articles/erosion/alps-erosion.tex
Discuss cristalline massif high erosion.

    For instance, low-elevation piedmont over-deepened basins such as Lake
    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore are
    here modelled to be the product of extensive Last Glacial Maximum-like
    glaciations. Other modelling work suggests that features such as these
    likely form when water can access the glacier bed and encourage glacier
    sliding and, thus, erosion \citep{Herman.etal.2011}. The model herein does
    not include subglacial hydrology, yet the role of climate warming to
    increasing glacier erosion to the lower elevations remains evident
    (Fig.~\ref{fig:hypsogram}).
    Alpine terrain with elevations above 1000\,m are modelled to
    experience virtually no erosion from 35 to 18\,ka BP
    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
    intermediate-elevation glacial valleys of the Alps as Lauterbrunnental
    (Fig.~\ref{fig:landscape}), Haslital (Upper Aare) or Bout du Monde (Upper
    Giffre) would date from periods of intermediate glacier extension.
    The validity of the model results at high elevation are limited by the 1-km
    resolution, the absence of glacier cirque erosion processes
    \citep[cf.][]{Sanders.etal.2012}, and the shallow-ice glacier flow
    physics \citep{Imhof.etal.2019}. However,
    the lack of rapid erosion at high elevations for much
    of the modelled glacial cycle implies that the highest mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would have formed
    over interglacial periods, when glaciers are confined to high terrain
    \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
    Himalayan glaciers and may limit erosion for the high elevation portion of
    that mountain range during cold periods as well, failing to offset uplift
    rates \citep{Harper.Humphrey.2003}.

>>>>

    For instance, low-elevation piedmont over-deepened basins such as Lake
    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore are
    here modelled to be the product of extensive Last Glacial Maximum-like
    glaciations. Other modelling work suggests that features such as these
    likely form when water can access the glacier bed and encourage glacier
    sliding and, thus, erosion \citep{Herman.etal.2011}. The model herein does
    not include subglacial hydrology, yet the role of climate warming to
    increasing glacier erosion to the lower elevations remains evident
    (Fig.~\ref{fig:hypsogram}).
    Alpine terrain with elevations above 1000\,m are modelled to
    experience virtually no erosion from 35 to 18\,ka BP
    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
    intermediate-elevation glacial valleys of the Alps as Lauterbrunnental
    (Fig.~\ref{fig:landscape}), Haslital (Upper Aare) or Bout du Monde (Upper
    Giffre) would date from periods of intermediate glacier extension.

    The validity of the model results at high elevation is discussable.
    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
    Ötztal and Tauern Massifs locally exhibit a strikingly high erosion
    potential. However, the computation of glacier flow velocities on such
    steep surfaces is strongly limited by the model horizontal resolution of
    1\,km, the shallow-ice glacier flow physics \citep{Imhof.etal.2019}, and
    PISM's current mass-conservation heuristics \citep{Imhof.2021}. Besides,
    bergschrund (rimaye) processes likely to dominate interglacial cirque
    erosion at such altitudes \citep{Sanders.etal.2012} are not captured by the
    velocity-based glacier erosion power-laws. Nevertheless,
    the lack of rapid erosion at high elevations for much
    of the modelled glacial cycle implies that the highest mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would have formed
    over interglacial periods, when glaciers are confined to high terrain
    \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
    Himalayan glaciers and may limit erosion for the high elevation portion of
    that mountain range during cold periods as well, failing to offset uplift
    rates \citep{Harper.Humphrey.2003}.
